---
title: "lab_02"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? **A1. Atmospheric and Oceanic Science**

```{r}
climate_courses <- umd_courses |> filter(str_detect(title, "Climate") & seats >=1)|>
  group_by(department) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(count_classes))
climate_courses

```

**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why?

**A2: University honors is interesting because it has the third most, even more than geographical sciences which seems a little crazy- but I am assuming since it covers a lot of bases across the entire school that is why.**

**Q3** Which departments offered the most courses in the Fall 2024 term? And which departments in that term have at least one syllabus posted? Answer both of those questions and then compare the top five departments - how are they different?

**A3. The departments that offered the most courses are business and management(174), school of music(158), info studies,** english, and teachhing and learning policy and leadership**. Business and Management has 106 courses with a syllabus posted and economics then jumps up to second, while school of music has 52 courses with a syllabus. Info sciences is still in the top 5 but now engineering has popped into the 5th place. But for having over 100 courses for the fall, the number is cut in half for most of them when you factor in the syllabus.**

```{r}
fall_2024_courses <- umd_courses |> filter(term == "202408") |> filter(syllabus_count >= 1) |>
group_by(department) |>
  summarise(
    count_courses = n()
  ) |>
  arrange(desc(count_courses))
fall_2024_courses


```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as a new dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name, and we're not looking in payee_name). Write a sentence describing the pattern of spending over time.

**A4. Moore's campaign spending went up throughout each month- and saw a lot of increases in June and October of 2021- which lines up with when e won the party primary**

```{r}
update_maryland_expenses <- maryland_expenses|> mutate(expenditure_date=mdy(expenditure_date))

update_maryland_expenses
```

```{r}
  to_moore_comittee_by_month <- maryland_expenses_with_month |> 
    filter(str_detect(committee_name, "Moore"))  |>
  group_by(`month)  |>
  summarise(total_amount = sum(amount)) |> arrange(desc(month))
to_moore_comittee_by_month

```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for all others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe). **A5**

```{r}
maryland_expenses_by_state <- maryland_expenses |> mutate(spending_location = case_when(str_detect(address, "Maryland") - "In State"                             .default = "Out of State")) |>
  group_by(spending_location) |> summarise(amount_spent = sum(amount))
maryland_expenses_by_state

```

**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab. **A6.**

```{r}
```
